The hydra and the rooster

### 2020-08-06 latex/chapter-ks.tex
correction on pdf file

Our proofs are  constructive and require no axioms: they are  closed terms of the CIC, and are mainly composed on function definitions and proofs of properties of these functions. 
They  share much theoretical material with Kirby and Paris', although they do not use any knowledge about Peano arithmetic nor model  theory.  The combinatorial arguments we use and implement
come from 
 an article by J.~Ketonen and R.~Solovay~\cite{KS81}, already  cited in the work
 by L.~Kirby et J.~Paris on the termination of Goodstein sequences and hydra battles~\cite{KP82}.
 Section $2$ of this article: ''A hierarchy of probably recursive functions'', contains a systematic study of \emph{canonical sequences}, which are closely related to
rounds of hydra battles.

@@

\paragraph{Remark}
In~\cite{KS81}, canonical sequences are defined for any ordinal $\alpha <\epsilon_0$,
by stating that if $\alpha$ is a successor ordinal, the sequence associated with 
$\alpha$ is simply the constant sequence whose terms are equal to the predecessor of $\alpha$. Likewise, we define the canonical sequence of $0$ as the
sequence whose all terms are equal to $0$.

This convention allows us to make total the function that maps any ordinal $\alpha$ and natural number $i$ to the $i$-th item of the canonical sequence associated with $\alpha$.

First, let us recall how canonical sequences are defined in~\cite{KS81}. For efficiency's sake, we decided not to implement directly K.\&S's definitions, but to define in \gallina{} simply typed structurally recursive functions which share the abstract properties which are used in the mathematical proofs\footnote{With a small diffrence: the $0$-th term of the canonical sequence is not the same in our development as in~\cite{KS81}.}

@@

Our definition may look more complex than the mathematical one, but
uses plain structural recursion over the type \coqsimple{T1}. Thus, tactics like
\coqsimple{cbn}, \coqsimple{simpl}, \coqsimple{compute}, etc., are applicable. For simplicity's sake, 
we used an auxiliary function \texttt{canonS} of type \texttt{T1 -> nat  -> T1} such that
\texttt{canonS  $\alpha$ $i$} is equal to $\canonseq{\alpha}{i+1}$.

@@

\subsection{Definition}

We shall denote by $\alpha\xrightarrow [s]{}\beta$
  the proposition 
``there is a path from $\alpha$ to $\beta$ with $s$ as the sequence of indices
$s_0,s_1,\dots,s_{l-1}$ ''.

For instance, we have $\omega*2 \xrightarrow[2,2,2]{}\omega$, through the 
sequence of ordinals $\omega\times 2, \omega+2,\omega+1,\omega$.

@@

Indeed, if $\alpha$ is a limit ordinal, the first element of $s$ can be any integer $i$ such that $\beta<\canonseq{\alpha}(i)$, and if $\alpha$ is a successor ordinal,
then the sequence $s$ can start with any positive integer.

For instance, we have also 
$\omega*2 \xrightarrow[3,4,5,6]{}\omega$. 
Likewise,
$\omega*2 \xrightarrow[{[3,14]}]{} 0$ and
$\omega*2 \xrightarrow[s] 0$, where $s$ is the sequence containing eight times the number $3$.

@@

On may ask whether the preceding results are not too restrictive, since they 
refer to a particular data type \texttt{T1}.
In fact, our representation of ordinals strictly less than 
 $\epsilon_0$ is faithful to their mathematical definition, at least 
Kurt Schütte's~\cite{schutte}, as proved in Chapter~\vref{chap:schutte}.
(please see also \href{../src/html/hydras.Schutte.Injection_E0.html}{the module \texttt{Ordinals.Schutte.Injection\_E0}}).

>>>>

Our proofs are  constructive and require no axioms: they are  closed terms of the CIC, and are mainly composed on function definitions and proofs of properties of these functions. 
They  share much theoretical material with Kirby and Paris', although they do not use any knowledge about Peano arithmetic nor model  theory.  The combinatorial arguments we use and implement
come from 
 an article by J.~Ketonen and R.~Solovay~\cite{KS81}, already  cited in the work
 by L.~Kirby et J.~Paris.% on the termination of Goodstein sequences and hydra battles~\cite{KP82}.
 Section $2$ of this article: ''A hierarchy of probably recursive functions'', contains a systematic study of \emph{canonical sequences}, which are closely related to
rounds of hydra battles.

@@

\begin{remark}
  In~\cite{KS81}, canonical sequences are defined for any ordinal $\alpha <\epsilon_0$,
by stating that if $\alpha$ is a successor ordinal $\beta+1$,  the sequence associated with 
$\alpha$ is simply the constant sequence whose terms are equal to $\beta$.
Likewise, we define the canonical sequence of $0$ as the 
sequence all terms of which are equal to $0$.

This convention allows us to make total the function that maps any ordinal $\alpha$ and natural number $i$ to the ordinal $\canonseq{\alpha}{i}$.

\end{remark}

First, let us recall how canonical sequences are defined in~\cite{KS81}. For efficiency's sake, we decided not to implement directly K.\&S's definitions, but to define in \gallina{} simply typed structurally recursive functions which share the abstract properties which are used in the mathematical proofs\footnote{With a small diffrence: the $0$-th term of the canonical sequence is not the same in our development as in~\cite{KS81}.}.

@@

Our definition may look more complex than the mathematical one, but
uses plain structural recursion over the type \coqsimple{T1}. Thus, tactics like
\coqsimple{cbn}, \coqsimple{simpl}, \coqsimple{compute}, etc., are applicable. For simplicity's sake, 
we used an auxiliary function \texttt{canonS} of type \texttt{T1 -> nat  -> T1} such that
(\texttt{canonS  $\alpha$ $i$}) is equal to $\canonseq{\alpha}{i+1}$.

@@

Let $s$ be the sequence $\langle s_0,s_1,\dots, s_{l-1} \rangle$. We describe the
existence of such a path with the notation $\alpha\xrightarrow [s]{}\beta$.

For instance, we have $\omega*2 \xrightarrow[2,2,2]{}\omega$, through the 
path $\omega\times 2, \omega+2,\omega+1,\omega$.

\begin{remark}

@@

Indeed, if $\alpha$ is a limit ordinal, the first element of $s$ can be any integer $i$ such that $\beta<\canonseq{\alpha}{i}$, and if $\alpha$ is a successor ordinal,
then the sequence $s$ can start with any positive integer.

For instance, we have also 
$\omega*2 \xrightarrow[3,4,5,6]{}\omega$. 
Likewise,
$\omega*2 \xrightarrow[{[3,14]}]{} 0$ and
$\omega*2 \xrightarrow[3,3,3,3,3,3,3,3]{} 0$.
\end{remark}

\subsection{Formal definition}

@@

On may ask whether the preceding results are not too restrictive, since they 
refer to a particular data type \texttt{T1}.
In fact, our representation of ordinals strictly less than 
 $\epsilon_0$ is faithful to their mathematical definition, at least 
Kurt Schütte's~\cite{schutte}, as proved in Chapter~\vref{chap:schutte}.
(please see also \href{../src/html/hydras.Schutte.Correctness_E0.html}{the module \texttt{Ordinals.Schutte.Correctness\_E0}}).

### 2020-08-06 latex/chapter-termination.tex
correction on pdf file

Another way is encapsulate conditions of the form \texttt{(nf $\alpha$)} in
often used predicates. For instance, we introduced the restriction of \texttt{lt} to terms in normal form, and provided a handy notation for this restriction.

@@

The type \texttt{T1} is not a correct ordinal notation, since it contains terms that are not in Cantor normal form. In order to define a type whose inhabitants represent just ordinals, we build a type gathering a term of type \texttt{T1} and a proof that this term is in normal form.

@@

For upgrading constants and fonctions of \texttt{T1}, we have to prove that 
the term they build is in normal form.
For instance, let us represent the ordinals $0$ $\omega$ as instances of the class \texttt{E0}.

@@

Nevertheless, we built an instance of \texttt{ON}, and proved its correction w.r.t. Schutte's model.

@@

\begin{Coqsrc}
Fixpoint succ (alpha:T1) : T1 :=
  match alpha with zero => 1
             | ocons zero n _ => ocons zero (S n) zero
             | ocons beta n gamma => ocons beta n (succ gamma)
 end.
\end{Coqsrc}

@@

The function \texttt{pp: T1 -> ppT1} converts any closed term of type \texttt{T1} into a human-readable expression. For instance, let us convert the term \texttt{alpha\_0} of~\pageref{alpha0-def}.

@@

@@

First, we prove by transfinite induction on $\alpha$ that the restriction of $\oplus$ to the
interval $[0..\alpha[$ is commutative.

@@

Then, we infer  $\oplus$'s commutativity for any pair of ordinals:
Let $\alpha$ and $\beta$ be two ordinals strictly less than $\epsilon_0$. Both ordinals $\alpha$ and $\beta$ are
strictly less than $\textrm{max}(\alpha,\beta)+1$.

    Thus, we have just to apply the lemma \coqsimple{oplus\_comm\_0}.

@@

\begin{Coqsrc}
Theorem m_nf : forall h, nf (m h).
Proof.
 intro h; elim h using Hydra_rect2 
            with (P0 := fun s =>  nf (ms s)).
 (* ... *)

Theorem ms_nf : forall s, nf (ms s).
Proof with auto with T1.
(* ... *)
\end{Coqsrc}

@@

Let us recall three results we have proved so far.
\begin{itemize}
\item There exists a strictly decreasing variant mapping \texttt{Hydra} into 
$[0,\epsilon_0[$ for proving the termination of any hydra battle
\item There exists \emph{no} such variant from \texttt{Hydra} into $[0,\omega^2[$ and
a fortiori into $[0,\omega[$.
\end{itemize}

>>>>

One may encapsulate conditions of the form \texttt{(nf $\alpha$)} in
the most used predicates. For instance, we introduce the restriction of \texttt{lt} to terms in normal form, and provide a handy notation for this restriction.

@@

\subsubsection{A sigma type for \texorpdfstring{$\epsilon_0$}{epsilon0}}

The type \texttt{T1} is not a correct ordinal notation, since it contains terms that are not in Cantor normal form. In order to define a type whose inhabitants represent just ordinals, we build a type gathering a term of type \texttt{T1} and a proof that this term is in normal form.
This representation will be used in section~\ref{sect:equations} in order to define recursive functions with the plug-in \texttt{coq-equations}, and also to build an instance of the class \texttt{ON} of ordinal notations.

@@

For upgrading constants and fonctions of \texttt{T1}, we have to prove that 
the term they build is in normal form.
For instance, let us represent the ordinals $0$ and $\omega$ as instances of the class \texttt{E0}.

@@

@@

\begin{Coqsrc}
Fixpoint succ (alpha:T1) : T1 :=
  match alpha with 
   | zero => 1
   | ocons zero n _ => ocons zero (S n) zero
   | ocons beta n gamma => ocons beta n (succ gamma)
 end.
\end{Coqsrc}

@@

The function \texttt{pp: T1 -> ppT1} converts any closed term of type \texttt{T1} into a human-readable expression. For instance, let us convert the term \texttt{alpha\_0}.

@@

We built an instance of \texttt{ON}, and proved its correction w.r.t. Schutte's model.

\index{Projects}
\begin{project}
 \emph{This exercise is a continuation of Project~\vref{exo:ON-mult}.}
Define an ordinal notation \texttt{Omega2} for $\omega^2=\omega\times\omega$ (using
\texttt{ON\_mult}).

Prove that \texttt{Omega2} is a sub-notation of \texttt{Epsilon0}.

Define on \texttt{Omega2} an addition compatible with the addition on \texttt{Epsilon0}.

\textbf{Hint}. You may use the following definition (in 
\href{../src/html/hydras.OrdinalNotations.Definitions.html}{OrdinalNotations.Definitions}).

\begin{Coqsrc}
Definition SubON_same_op  `{OA : @ON A ltA  compareA}
       `{OB : @ON B ltB  compareB}
       {iota : A -> B} 
       {alpha: B}
       {_ : SubON OA OB alpha iota}
       (f : A -> A -> A)
       (g : B -> B -> B)
  :=
  forall x y,  iota (f x y) = g (iota x) (iota y).
\end{Coqsrc}

@@

First, we prove by transfinite induction on $\alpha$ that the restriction of $\oplus$ to the
interval $[0..\alpha)$ is commutative.

@@

Then, we infer  $\oplus$'s commutativity for any pair of ordinals:
Let $\alpha$ and $\beta$ be two ordinals strictly less than $\epsilon_0$. Both ordinals $\alpha$ and $\beta$ are
strictly less than $\textrm{max}(\alpha,\beta)+1$.
    Thus, we have just to apply the lemma \coqsimple{oplus\_comm\_0}.

@@

\begin{Coqsrc}
Lemma m_nf : forall h, nf (m h).
Proof.
 intro h; elim h using Hydra_rect2 
            with (P0 := fun s =>  nf (ms s)).
 (* ... *)

Lemma ms_nf : forall s, nf (ms s).
Proof with auto with T1.
(* ... *)
\end{Coqsrc}

@@

Let us recall three results we have proved so far.
\begin{itemize}
\item There exists a strictly decreasing variant which maps \texttt{Hydra} into 
the segment $[0,\epsilon_0)$ for proving the termination of any hydra battle
\item There exists \emph{no} such variant from \texttt{Hydra} into $[0,\omega^2)$ and
a fortiori into $[0,\omega)$.
\end{itemize}
